Modelling the innovation and extinction of archaeological ideas

### 2018-01-27 analysis/paper/paper.Rmd
add title and authors

---
title: "Title Goes Here"
author:
  - author 1
  - author 2
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: journal-of-archaeological-science.csl
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

Here is a citation [@Marwick2017]

@@

@@

We have `r hm_items` articles in our study. The publication dates range from `r time_range[1]` to  `r time_range[2]`. The bibliographic data for these articles was obtained from a search of Clarivate Analytics Web of Science, using the keyword 'archaeolog' and filtering the results to keep only research articles.

The distribution of articles is strongly skewed to more recent years, with a rapid increase after the mid-2000s. For the rest of our analysis we're going to limit the time frame to 1975-207 because the data from beofre 1975 is incomplete for some key journals, such as _American Antiquity_.

@@

library(scales)
ggplot(xxx,
       aes(year,
            n)) +
  geom_col() +
  scale_y_log10(breaks = trans_breaks('log10', function(x) 10^x),
                  labels = trans_format('log10', math_format(10^.x))) +
  theme_bw() +
  ggtitle("Citations per year in archaeology articles (since 1975)")
```

@@

```{r}
most_cited <- 
  xx %>% 
  filter(year >= 1975) %>% 
  count(.out, sort = TRUE)  %>% 
  slice(1:10)
```

@@

most_cited_per_decade <- 
  my_df %>% 
  filter(year >= 1975) %>% 
  group_by(decade) %>% 
  count(.out, sort = TRUE) %>% 
  filter(n == max(n))
```

@@

# age distribution of "Archaeology as Anthropology" citations, as a proportion
# of all citations in each year
binford_1962_cites_per_year <-
  xx %>%
  filter(year >= 1975) %>% 
  filter(grepl("BINFORD LR 1962 AM ANTIQUITY V28 P217", 
               .out)) %>%
    group_by(year) %>%
    summarise(count = n()) %>%
    left_join(xxx, 
              by = "year") %>%
    mutate(prop = count / n) %>%
    distinct(year, 
             prop, 
             n)

@@

Data sets in occurrence format, such as we have here, provide the opportunity to incoporate uncertainty about the preservation of the observations when analyzing origination and extinction rates. If a researcher does not have a high level of certainty that the dates of first and last occurrence are correct, such as is often the case with fossils or artifacts,  preservation can be informative [@silvestro2014].

@@

We can make a histogram showing the distribution of the number of years each cited article was cited for (artcle lifespan):

@@

refs_occurance_format_hist_year %>% 
  mutate(end_year = year + number_of_years_cited) %>% 
  ungroup() %>% 
  select(Species, year, end_year) %>% 
  ggplot() +
  geom_linerange(aes(reorder(x = Species, year),
                     ymin = year,
                     ymax = end_year)) +
    coord_flip() +
  ylim(c(1975, 
         2019))

@@

@@

```{r, eval=TRUE}
source("pyrate_mcmc_logs/lineage_pyrate_data_0_BD_marginal_rates_RTT.r")
age_rev <-
  2017 + age # adjust for the most recent date in each of the datasets

>>>>

---
title: "Modelling the innovation and extinction of archaeological ideas"
author:
  - Ben Marwick, University of Washington
  - Erik Gjesfjeld, University of Cambridge 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: journal-of-archaeological-science.csl
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

The history of archaeology is often told as a sequence of prominent individuals and their publications. Due to the focus on big names and big papers, the diversity of archaeological publications is often underestimated. Here we introduce a quantitative method that illuminates historical trends in archaeological writing by investigating a large number of journal articles. We use a Bayesian framework developed for estimating speciation, extinction, and preservation rates from incomplete fossil data. We model archaeological ideas within this framework by equating citations of archaeological literature to occurrences in the fossil record. We obtained reference lists for 12,000 journal articles published between 1975 and 2017 and explored the chronological distribution of cited papers to identify periods of innovation and extinction. We discuss how our modeling approach helps to quantify the diversification of archaeological publications and our broader understanding about the history of archaeological thought.

@@

# how many 1975-2017
hm_items_1975 <- items_wth_refs %>% filter(year >= 1975) %>% nrow

@@

We have `r prettyNum(hm_items, big.mark = ",")` articles in our study. The publication dates range from `r time_range[1]` to  `r time_range[2]`. The bibliographic data for these articles was obtained from a search of Clarivate Analytics Web of Science, using the keyword 'archaeolog' and filtering the results to keep only research articles.

The distribution of articles is strongly skewed to more recent years, with a rapid increase after the mid-2000s. For the rest of our analysis we're going to limit the time frame to the `r prettyNum(hm_items_1975, big.mark = ",")` published during 1975-2017 because the data from before 1975 is incomplete for some key journals, such as _American Antiquity_.

@@

library(scales)
ggplot(xxx,
       aes(year,
            n)) +
  geom_col() +
  #scale_y_log10(breaks = trans_breaks('log10', function(x) 10^x),
  #                labels = trans_format('log10', math_format(10^.x))) +
  theme_bw() +
  ggtitle("Citations per year in archaeology articles (since 1975)")
```

@@

Here are the top ten most cited items overall:

```{r}
most_cited <- 
  xx %>% 
  filter(year >= 1975) %>% 
  count(.out, sort = TRUE)  %>% 
  slice(1:10)

knitr::kable(most_cited)
```

And here are the most cited items per decade (there is a tie for the 1970-1980 data):

@@

most_cited_per_decade <- 
  my_df %>% 
  filter(year >= 1975) %>% 
  group_by(decade) %>% 
  count(.out, sort = TRUE) %>% 
  filter(n == max(n))

knitr::kable(most_cited_per_decade)
```

@@

```{r}
# age distribution of "Archaeology as Anthropology" citations, as a proportion
# of all citations in each year
binford_1962_cites_per_year <-
  xx %>%
  filter(year >= 1975) %>% 
  filter(grepl("BINFORD LR 1962 AM ANTIQUITY V28 P217", 
               .out)) %>%
    group_by(year) %>%
    summarise(count = n()) %>%
    left_join(xxx, 
              by = "year") %>%
    mutate(prop = count / n) %>%
    distinct(year, 
             prop, 
             n)

@@

Data sets in occurrence format, such as we have here, provide the opportunity to incoporate uncertainty about the preservation of the observations when analyzing origination and extinction rates. If a researcher does not have a high level of certainty that the dates of first and last occurrence are correct, such as is often the case with fossils or artifacts,  preservation can be informative.

@@

We can make a histogram showing the distribution of the number of years each cited article was cited for (article lifespan):

@@

refs_occurance_format_hist_year %>% 
  mutate(titles  = str_wrap(str_trunc(Species, 40), 10)) %>% 
  mutate(end_year = year + number_of_years_cited) %>% 
  mutate(end_year = if_else(end_year > 2017, 2017, end_year)) %>% 
  ungroup() %>% 
  select(titles, year, end_year) %>% 
  ggplot() +
  geom_linerange(aes(x = reorder(titles, year),
                     ymin = year,
                     ymax = end_year),
                 size = 2) +
  geom_point(aes(x = reorder(titles, year),
                 year),
             size = 4) +
    geom_point(aes(x = reorder(titles, year),
                 end_year),
               size = 4) +
  ylim(c(1975, 
         2017)) +
  coord_flip()

@@

```{r, eval=FALSE}

@@

```{r, eval=TRUE}
source(paste0(here::here(), "/pyrate_mcmc_logs/lineage_pyrate_data_0_BD_marginal_rates_RTT.r"))

age_rev <-
  2017 + age # adjust for the most recent date in each of the datasets
